Supplementary Note 1: Metrics to compare two trajectories

### 2018-08-27 manuscript/supplementary_notes/metrics/metrics.Rmd
update metrics

experiment("02-metric_characterisation/01-metric_conformity")

refs <- setup_refs()
figs <- setup_figs()
```

@@

All metrics described here were implemented within the dyneval R package ([https://github.com/dynverse/dyneval](https://github.com/dynverse/dyneval)).

@@

## $`r label_metric("isomorphic", "latex")`$, $`r label_metric("edge_flip", "latex")`$ and $`r label_metric("him", "latex")`$: Edit distance between two trajectory topologies

@@

The $`r label_metric("isomorphic", "latex")`$ score returns 1 if two graphs are isomorphic, and 0 if they were not. For this, we used the used the BLISS algorithm [@junttila_engineeringefficientcanonical_2007], as implemented in the R *igraph* package.

The $`r label_metric("edge_flip", "latex")`$ score was defined as the minimal number of edges which should be added or removed to convert one network into the other, divided by the total number of edges in both networks. This problem is equivalent to the maximum common edge subgraph problem, a known NP-hard problem without a scalable solution [@bahiense_maximumcommonedge_2012]. We implemented a branch and bound approach for this problem, using several heuristics to speed up the search:

@@

The $`r label_metric("him", "latex")`$ metric (Hamming-Ipsen-Mikhailov distance) [@jurmanHIMGlocalMetric2015] which was adopted from the R nettools package [https://github.com/filosi/nettools](https://github.com/filosi/nettools). It uses an adjacency matrix which was weighted according to the lengths of each edges within the milestone network. Conceptually, $`r label_metric("him", "latex")`$ is a linear combination of:

@@

We compared the three scores for several common topologies (`r ref("fig", "topology_scores_overview")`). While conceptually very different, the $`r label_metric("edge_flip", "latex")`$ and $`r label_metric("him", "latex")`$ still produce similar scores (`r ref("fig", "topology_scores_overview", "b")`). The $`r label_metric("him", "latex")`$ tends to punish the detection of cycles, while the $`r label_metric("edge_flip", "latex")`$ is more harsh for differences in the number of bifurcations (`r ref("fig", "topology_scores_overview", "c")`). The main difference however is that the $`r label_metric("him", "latex")`$ takes into account edge lengths when comparing two trajectories, as illustrated in (`r ref("fig", "topology_scores_overview", "d")`).

```{r, results = "asis"}
add_fig(
  result_file("topology_scores_overview.rds", experiment = "02-metric_characterisation/02-metric_examples"),
  "topology_scores_overview",
  glue::glue("${label_metric('isomorphic', 'latex')}$, ${label_metric('edge_flip', 'latex')}$ and ${label_metric('him', 'latex')}$ ."),
  "",
  width = 12,
  height = 12
)
```

@@

- If the two trajectories should only be compared when the topology is exactly the same, the $`r label_metric("isomorphic", "latex")`$ should be used.
- If it is important that the topologies are similar (but possible exactly isomorphic), the $`r label_metric("edge_flip", "latex")`$ is most appropriate.
- If the topologies should be similar, but shorter edges should not be punished as hard as longer edges, the $`r label_metric("HIM", "latex")`$ is most appropriate.

## $`r label_metric("F1_branches", "latex")`$ and $`r label_metrics("F1_milestones", "latex")`$: Comparing how well the cells are clustered in the trajectory

@@

```{r, results = "asis"}
add_fig(
  result_file("clustering_scores_overview.rds", experiment = "02-metric_characterisation/02-individual_metrics"),
  "clustering_scores_overview",
  glue::glue("Mapping cells to their closest milestone or branch for the calculation of the ${label_metric('F1_branches', 'latex')}$, ${label_metric('F1_edges', 'latex')}$ and ${label_metric('F1_branches', 'latex')}$ ."),
  "",
  width = 12,
  height = 5
)
```

## $`r label_metric("correlation", "latex")`$: Correlation between geodesic distances

When the position of a cell is the same in both the gold standard and the prediction, its distances to all other cells in the trajectory should also be the same. This observation is the basis for the $`r label_metric("correlation", "latex")`$ metric.

@@

Calculating the pairwise distances between cells scales quadratically with the number of cells, and would therefore not be scaleable for large datasets. For this reason, a set of waypoint cells are defined *a priori*, and only the distances between the waypoint cells and all other cells is calculated, in order to calculate the correlation of geodesic distances of two trajectories (`r ref("fig", "waypoints_overview", "a")`). These cell waypoints are determined by viewing each milestone, edge between two milestones and region of delayed commitment as a collection of cells. We do stratified sampling from each collection of cells by weighing them by the total number of cells within that collection. For calculating the $`r label_metric("correlation", "latex")`$ between two trajectories, the distances between all cells and the union of both waypoint sets is computed.

To select the number of cell waypoints, we need to find a trade-off between the accuracy versus the time to calculate $`r label_metric("correlation", "latex")`$. To select an optimal number of cell waypoints, we used the synthetic dataset with the most complex topology, and determined the $`r label_metric("correlation", "latex")`$ at different levels of both cell shuffling and number of cell waypoints (`r ref("fig", "waypoints_overview", "b")`). We found that using cell waypoints does not induce a systematic bias in the $`r label_metric("correlation", "latex")`$, and that its variability was relatively minimal when compared to the variability between different levels of cell shuffling when using 100 or more cell waypoints.

```{r, results = "asis"}
add_fig(
  result_file("waypoints_overview.rds", experiment = "02-metric_characterisation/02-individual_metrics"),
  "waypoints_overview",
  glue::glue("Determination of cell waypoints"),
  glue::glue("a) Illustration of the stratified cell sampling using an example dataset (top). Each milestone, edge between two milestones and region of delayed commitment is seen as a collection of cells (middle), and the number of waypoints (100 in this case) are divided over each of these collection of cells (bottom). b) Accuracy versus time to calculate ${label_metric('correlation', 'latex')}$"),
  width = 12, height = 8
)
```

## $`r label_metric("rf_nmse", "latex")`$ and $`r label_metric("lm_nmse", "latex")`$: Using the positions of the cells within one trajectory to predict the cellular positions in the other trajectory

@@

Specifically, we implemented two metrics which predict the milestone percentages from the gold standards by using the predicted milestone percentages as features  (`r ref("fig", "metrics_prediction")`). We did this with two regression methods, linear regression ($\textit{lm}$, using the R `lm` function) and Random Forest ($\textit{rf}$, implemented in the *ranger* package [@wright_rangerfastimplementation_2017]). In both cases, the accuracy of the prediction was measured using the Mean Squared error ($\mathit{MSE}$), in the case of Random forest we used the out-of-bag mean-squared error. Next, we calculated $\mathit{MSE}_{worst}$ equal to the $\mathit{MSE}$ when predicting all milestone percentages as the average. We used this to calculate the normalised mean squared error as $\mathit{NMSE} = 1 - \frac{\mathit{MSE}}{\mathit{MSE}_{worst}}$. We created a regression model for every milestone in the gold standard, and averaged the $\mathit{NMSE}$ values to finally obtain the $`r label_metric("rf_nmse", "latex")`$ and $`r label_metric("lm_nmse", "latex")`$ scores.

```{r, results = "asis"}
add_fig(
  result_file("metrics_prediction.svg", "manual_figures"),
  "metrics_prediction",
  glue::glue("The calculation of ${label_metric('lm_nmse', 'latex')}$ distances on a small example trajectory."),
  ""
)
```

@@

## $`r label_metric("featureimp_cor", "latex")`$: The accuracy of dynamical differentially expressed features/genes.

Perhaps the main advantages of studying cellular dynamic processes using single-cell -omics data is that the dynamics of gene expression can be studied for the whole transcriptome. This allows the study of dynamic regulatory networks and gene expression modules. Such analyses rely on a "good-enough" cellular ordering, so that the cellular ordering can be used to identify dynamical differentially expressed genes.

To calculate the $`r label_metric("featureimp_cor", "latex")`$ we used Random forest regression to rank all the features according to their importance in predicting the positions of cells in the trajectory. Specifically, we calculated the geodesic distances for each cell to all milestones in the trajectory. We then trained a Random Forest regression model (implemented in the R _ranger_ package [@wright_rangerfastimplementation_2017], [https://github.com/imbs-hl/ranger](https://github.com/imbs-hl/ranger)) to predict these distances for each milestone, based on the expression of genes within each cell. We then extracted feature importances using the Mean Decrease in Impurity (`importance = 'impurity'` parameter of the `ranger` function), as illustrated in (`r ref("fig", "featureimp_overview.rds")`). The overall importance of a feature (gene) was then equal to the mean importance over all milestones.

@@

Random forest regression has two main hyperparameters. The number of trees to be fitted (`num_trees` parameter) was fixed to `10000` to provide accurate estimates of the feature importance. The number of features on which can be split (`mtry` parameter) was set to 1% of all available features (instead of the default square-root of the number of features), as to make sure that predictive but highly correlated feautres (omnipresent in transcriptomics data) can all get higher on the ranking.

@@

@@

```{r}
assessments %>%
  left_join(rules %>% select(id, name), by = c("rule_id" = "id")) %>%
  unnest(conformity) %>%
  mutate(
    metric_id = label_metrics(metric_id, label_type = "latex") %>% paste0("$", ., "$") %>% forcats::fct_inorder(),
    conforms = kableExtra::cell_spec(
      ifelse(conforms, "\U2714", "\U2716"),
      background = ifelse(conforms, "#2ECC40", "#FF4136"),
      format = "latex"
    ),
    rule_id = forcats::fct_inorder(rule_id)
  ) %>%
  spread(metric_id, conforms) %>%
  select(-rule_id) %>% 
  knitr::kable(format = "latex", escape = FALSE, booktabs = TRUE) %>%
  kableExtra::kable_styling(full_width = FALSE) %>% 
  kableExtra::column_spec(1, width="15em") %>% 
  kableExtra::column_spec(2:20, width = "3em") %>% 
  kableExtra::row_spec(0, angle = 30) %>% 
  gsub("\\\\addlinespace", "", .)
```

@@

walkdf(assessments, function(assessment) {
  cat(knitr::knit_child("assessment.Rmd", quiet=TRUE, envir = environment()))
})
```

# Metrics used in the evaluation

## $`r label_metric("harm_mean", "latex")`$ The harmonic mean

>>>>

refs <- setup_refs()
figs <- setup_figs()
tables <- setup_tables()
```

@@

All metrics described here were implemented within the _dyneval_ R package ([https://github.com/dynverse/dyneval](https://github.com/dynverse/dyneval)).

@@

## `r label_metric("isomorphic")`, `r label_metric("edge_flip")` and `r label_metric("him")`: Edit distance between two trajectory topologies

@@

The `r label_metric("isomorphic")` score returns 1 if two graphs are isomorphic, and 0 if they were not. For this, we used the used the BLISS algorithm [@junttila_engineeringefficientcanonical_2007], as implemented in the R *igraph* package.

The `r label_metric("edge_flip")` score was defined as the minimal number of edges which should be added or removed to convert one network into the other, divided by the total number of edges in both networks. This problem is equivalent to the maximum common edge subgraph problem, a known NP-hard problem without a scalable solution [@bahiense_maximumcommonedge_2012]. We implemented a branch and bound approach for this problem, using several heuristics to speed up the search:

@@

The `r label_metric("him")` metric (Hamming-Ipsen-Mikhailov distance) [@jurmanHIMGlocalMetric2015] which was adopted from the R nettools package [https://github.com/filosi/nettools](https://github.com/filosi/nettools). It uses an adjacency matrix which was weighted according to the lengths of each edges within the milestone network. Conceptually, `r label_metric("him")` is a linear combination of:

@@

We compared the three scores for several common topologies (`r ref("fig", "topology_scores_overview")`). While conceptually very different, the `r label_metric("edge_flip")`r and `r label_metric("him")`r still produce similar scores (`r ref("fig", "topology_scores_overview", "b")`). The `r label_metric("him")`r tends to punish the detection of cycles, while the `r label_metric("edge_flip")`r is more harsh for differences in the number of bifurcations (`r ref("fig", "topology_scores_overview", "c")`). The main difference however is that the `r label_metric("him")`r takes into account edge lengths when comparing two trajectories, as illustrated in (`r ref("fig", "topology_scores_overview", "d")`).

```{r, results = "asis"}
add_fig(
  result_file("topology_scores_overview.rds", experiment = "02-metric_characterisation/02-metric_examples"),
  "topology_scores_overview",
  glue::glue("{label_metric('isomorphic')}, {label_metric('edge_flip')} and {label_metric('him')} ."),
  "",
  width = 12,
  height = 12
)
```

@@

- If the two trajectories should only be compared when the topology is exactly the same, the `r label_metric("isomorphic")` should be used.
- If it is important that the topologies are similar (but possible exactly isomorphic), the `r label_metric("edge_flip")` is most appropriate.
- If the topologies should be similar, but shorter edges should not be punished as hard as longer edges, the `r label_metric("HIM")` is most appropriate.

## `r label_metric("F1_branches")` and `r label_metric("F1_milestones")`: Comparing how well the cells are clustered in the trajectory

@@

```{r, results = "asis"}
add_fig(
  result_file("clustering_scores_overview.rds", experiment = "02-metric_characterisation/02-individual_metrics"),
  "clustering_scores_overview",
  glue::glue("Mapping cells to their closest milestone or branch for the calculation of the {label_metric('F1_milestones')} and {label_metric('F1_branches')} ."),
  "",
  width = 12,
  height = 5
)
```

## `r label_metric("correlation")`: Correlation between geodesic distances

When the position of a cell is the same in both the gold standard and the prediction, its distances to all other cells in the trajectory should also be the same. This observation is the basis for the `r label_metric("correlation")` metric.

@@

Calculating the pairwise distances between cells scales quadratically with the number of cells, and would therefore not be scaleable for large datasets. For this reason, a set of waypoint cells are defined *a priori*, and only the distances between the waypoint cells and all other cells is calculated, in order to calculate the correlation of geodesic distances of two trajectories (`r ref("fig", "waypoints_overview", "a")`). These cell waypoints are determined by viewing each milestone, edge between two milestones and region of delayed commitment as a collection of cells. We do stratified sampling from each collection of cells by weighing them by the total number of cells within that collection. For calculating the `r label_metric("correlation")` between two trajectories, the distances between all cells and the union of both waypoint sets is computed.

To select the number of cell waypoints, we need to find a trade-off between the accuracy versus the time to calculate `r label_metric("correlation")`. To select an optimal number of cell waypoints, we used the synthetic dataset with the most complex topology, and determined the `r label_metric("correlation")` at different levels of both cell shuffling and number of cell waypoints (`r ref("fig", "waypoints_overview", "b")`). We found that using cell waypoints does not induce a systematic bias in the `r label_metric("correlation")`, and that its variability was relatively minimal when compared to the variability between different levels of cell shuffling when using 100 or more cell waypoints.

```{r, results = "asis"}
add_fig(
  result_file("waypoints_overview.rds", experiment = "02-metric_characterisation/02-individual_metrics"),
  "waypoints_overview",
  glue::glue("Determination of cell waypoints"),
  glue::glue("a) Illustration of the stratified cell sampling using an example dataset (top). Each milestone, edge between two milestones and region of delayed commitment is seen as a collection of cells (middle), and the number of waypoints (100 in this case) are divided over each of these collection of cells (bottom). b) Accuracy versus time to calculate {label_metric('correlation')}"),
  width = 12, height = 8
)
```

## `r label_metric("rf_nmse")` and `r label_metric("lm_nmse")`: Using the positions of the cells within one trajectory to predict the cellular positions in the other trajectory

@@

Specifically, we implemented two metrics which predict the milestone percentages from the gold standards by using the predicted milestone percentages as features  (`r ref("fig", "metrics_prediction")`). We did this with two regression methods, linear regression ($\textit{lm}$, using the R `lm` function) and Random Forest ($\textit{rf}$, implemented in the *ranger* package [@wright_rangerfastimplementation_2017]). In both cases, the accuracy of the prediction was measured using the Mean Squared error ($\mathit{MSE}$), in the case of Random forest we used the out-of-bag mean-squared error. Next, we calculated $\mathit{MSE}_{worst}$ equal to the $\mathit{MSE}$ when predicting all milestone percentages as the average. We used this to calculate the normalised mean squared error as $\mathit{NMSE} = 1 - \frac{\mathit{MSE}}{\mathit{MSE}_{worst}}$. We created a regression model for every milestone in the gold standard, and averaged the $\mathit{NMSE}$ values to finally obtain the `r label_metric("rf_nmse")` and `r label_metric("lm_nmse")` scores.

```{r, results = "asis"}
add_fig(
  result_file("metrics_prediction.svg", "manual_figures"),
  "metrics_prediction",
  glue::glue("The calculation of {label_metric('lm_nmse')} distances on a small example trajectory."),
  ""
)
```

@@

## `r label_metric("featureimp_cor")`: The accuracy of dynamical differentially expressed features/genes.

Perhaps the main advantages of studying cellular dynamic processes using single-cell -omics data is that the dynamics of gene expression can be studied for the whole transcriptome. This can be used to construct other models such as dynamic regulatory networks and gene expression modules. Such analyses rely on a "good-enough" cellular ordering, so that it can be used to identify dynamical differentially expressed genes.

To calculate the `r label_metric("featureimp_cor")` we used Random forest regression to rank all the features according to their importance in predicting the positions of cells in the trajectory. Specifically, we calculated the geodesic distances for each cell to all milestones in the trajectory. Next, we trained a Random Forest regression model (implemented in the R _ranger_ package [@wright_rangerfastimplementation_2017], [https://github.com/imbs-hl/ranger](https://github.com/imbs-hl/ranger)) to predict these distances for each milestone, based on the expression of genes within each cell. We then extracted feature importances using the Mean Decrease in Impurity (`importance = 'impurity'` parameter of the `ranger` function), as illustrated in (`r ref("fig", "featureimp_overview.rds")`). The overall importance of a feature (gene) was then equal to the mean importance over all milestones.

@@

Random forest regression has two main hyperparameters. The number of trees to be fitted (`num_trees` parameter) was fixed to `10000` to provide accurate and stable estimates of the feature importance (`r ref("fig", "featureimp_cor_distributions")`). The number of features on which can be split (`mtry` parameter) was set to 1% of all available features (instead of the default square-root of the number of features), as to make sure that predictive but highly correlated feautres (omnipresent in transcriptomics data) are not suppressed in the ranking.

```{r, results = "asis"}
add_fig(
  result_file("feautreimp_cor_distributions.rds", experiment = "02-metric_characterisation/02-individual_metrics"),
  "feautreimp_cor_distributions",
  glue::glue(""),
  glue::glue(""),
  width = 12, height = 6
)
```

# Overall metrics

An overall metric should not only focus on

@@

```{r}
experiment("02-metric_characterisation/01-metric_conformity")

dataset_design <- read_rds(derived_file("dataset_design.rds"))
```

Although most metrics described in the previous section make sense intuitively, this does not necessarily mean that these metrics are robust and will generate reasonable results when used for benchmarking. This is because different methods and datasets will all lead to a varied set of trajectory models:

- Real datasets have all cells grouped onto milestones
- Some methods place all cells in a region of delayed commitment, others never generate a region of delayed commitment
- Some methods always return a linear trajectory, even if a bifurcation is present in the data
- Some methods filter cells

A good metric, especially a good overall metric, should work in all these circumstances. To test this, we designed a set of rules to which a good metric should conform, and assessed empirically whether a metric conforms to these rules.

We generated a panel of toy datasets (using our _dyntoy_ package (https://github.com/dynverse/dyntoy)) with all possible combinations of:

- \# cells: `r sort(unique(dataset_design$num_cells)) %>% label_vector()`
- \# features: 200
- topologies: linear, bifurcation, multifurcating, tree, cycle, connected graph and disconnected graph
- Whether cells are placed on the milestones (as in real data) or on the edges/regions of delayed commitment between the milestones (as in synthetic data)

We then perturbed the trajectories in these datasets in certain ways, and tested whether the scores follow an expected pattern. An overview of the conformity of every metric is first given in `r ref("table", "conformity_overview")`. The individual rules and metric behaviour are discussed more into detail after that.

@@

```{r, results="asis"}
table <- map(c("html", "latex", "markdown"), function(format) {
  table <- assessments %>%
    left_join(rules %>% select(id, name), by = c("rule_id" = "id")) %>%
    unnest(conformity) %>% 
    mutate(
      metric_id = label_metrics(metric_id) %>% as.character() %>% forcats::fct_inorder()
    )

  if (format != "markdown") {
    table <- table %>% 
      mutate(
        conforms = kableExtra::cell_spec(
          ifelse(conforms, "\U2714", "\U2716"),
          background = ifelse(conforms, "#2ECC40", "#FF4136"),
          format = format
        ),
        rule_id = forcats::fct_inorder(rule_id)
      ) 
  } else {
    table <- table %>% 
      mutate(
        conforms = ifelse(conforms, "\U2714", "\U2716")
      )
  }

  table <- table %>%
    spread(metric_id, conforms) %>%
    select(-rule_id)

  if (format != "markdown") {
    knitr::kable(table, format = format, escape = FALSE, booktabs = TRUE) %>% 
      kableExtra::kable_styling(full_width = FALSE) %>% 
      kableExtra::column_spec(1, width="15em") %>% 
      kableExtra::column_spec(2:ncol(table), width = "3em") %>% 
      kableExtra::row_spec(0, angle = 30) %>% 
      gsub("\\\\addlinespace", "", .)
  } else {
    knitr::kable(table, format = format)
  }
}) %>% set_names(c("html", "latex", "markdown"))

add_table(
  table,
  "conformity_overview",
  "Overview of whether a particular metric conforms to a particular rule"
)
```

@@

walkdf(assessments[1, ], function(assessment) {
  cat(knitr::knit_child("assessment.Rmd", quiet=TRUE, envir = environment()))
})
```
